---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
``` {r}
require(ggplot2)
require(RColorBrewer)
require(dplyr)
require(ggthemes)
library(scales)
library(knitr)
```

## Loading and preprocessing the data

``` {r} 
# Reading the data 
steps <- read.csv("activity.csv", header =TRUE)
# Converting date to Date
steps$date <- as.Date(steps$date)
# Converting the interval to HH:MM format
steps$interval <- 
  format(strptime(formatC(steps$interval, width = 4, format = "d", flag = "0"), format="%H%M"), format = "%H:%M")
```
## What is mean total number of steps taken per day?
```{r}
## total steps per day calculation
StepsDaily <- steps %>%  
          group_by(date) %>%
          summarise(total = sum(steps))
# Histogram of mean steps per day
ggplot(StepsDaily, aes(total)) + 
  geom_histogram(binwidth = 1000, fill = "#C8A2C8", col='white') +
  labs(x = "Total steps daily", 
       y = "Frequency", 
       title = "Steps daily")
```
```{r}
## Mean steps per day
mean(StepsDaily$total, na.rm = TRUE)
## Median of steps per day
median(StepsDaily$total, na.rm = TRUE)
```

## What is the average daily activity pattern?
```{r}
intSteps1 <- steps %>% 
  group_by(interval) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
#Make a time series plot showing the average daily activity pattern
ggplot(intSteps1, aes(as.POSIXct(interval, format = "%H:%M"), mean)) + 
  geom_line(col = "#C8A2C8") +
  scale_x_datetime(labels = date_format("%H:%M"), 
                   date_breaks = "4 hours") +
  labs(x = "5 min Intervals", 
       y = "No. of Steps", 
       title = "Avg steps by time of day")
```


#Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
# Maximum number of average steps
max(intSteps1$mean)
# Which occurs in this time interval:
intSteps1[which.max(intSteps1$mean), ]
```

## Imputing missing values
```{r}
# Values having NA 
sum(!complete.cases(steps))
# Missing data given filled
inputSteps <- steps %>%
    group_by(interval) %>%
    mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
```
#Make a new histogram of the total number of steps taken each day
```{r}
# Total number of steps each day - histogram (assume missing values as mean)
inputDailySteps <- inputSteps %>% 
  group_by(date) %>%
  summarise(total = sum(steps))
ggplot(inputDailySteps, aes(total)) + 
  geom_histogram(binwidth = 1000, 
                 fill = "#C8A2C8", col = 'white') +
  labs(x = "Total Steps Daily", 
       y = "Frequency", 
       title = "Daily Steps")
## Mean number of steps per day
mean(inputDailySteps$total, na.rm = TRUE)
## Median number of steps per day
median(inputDailySteps$total, na.rm = TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# weekdays/weekend variable
WSteps <- inputSteps %>%
  mutate(Wkday = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday" ))
# Turning this into Factor
WSteps <- WSteps %>%
  mutate( wkday = as.factor(Wkday) )
# Group by interval and weekkday (factor) and again calculating the mean number of steps
WIntSteps <- WSteps %>% 
  group_by(interval,Wkday) %>%
  summarise(mean = mean(steps, na.rm = TRUE))
# Create time series plot with faceting on the new wkday factor
ggplot(WIntSteps, aes(as.POSIXct(interval, format = "%H:%M"), mean, col = Wkday)) + 
  geom_line(show.legend = F) +
  facet_grid(rows = WIntSteps$wkday) +
  scale_x_datetime(labels = date_format("%H:%M"), 
                   date_breaks = "4 hours") +
  labs(x = "5 Minute Interval", 
       y = "No. of Steps", 
       title = "Avg. steps by time of day")
```